---
title: "Final Deliverable Stock-Market-Retirement"
output: html_document
author: Mohriz Murad, Rohan Rajesh, author3, author 4
date: "2025-02-27"
---

```{r}
library(tidyverse)
getwd()
# Read the datasets
stocks <- read_delim("../../Data/sp500-monthly.csv")
metrics <- read_delim("../../Data/Cost_of_Living_and_Income_Extended.csv")
cpi_data <- read_delim("../../Data/CPIAUCSL.csv")

# Convert Date column in S&P 500 dataset to Date type
stocks$Date <- as.Date(stocks$Date)
cpi_data$observation_date <- as.Date(cpi_data$observation_date)

# Extract year from Date column
stocks$Year <- format(stocks$Date, "%Y")
stocks$Year <- as.numeric(stocks$Year)
cpi_data$Year <- format(cpi_data$observation_date, "%Y")
cpi_data$Year <- as.numeric(cpi_data$Year)

# Compute the total yearly CPI ensuring one row per year
cpi_yearly <- cpi_data %>% 
  group_by(Year) %>% 
  summarise(Avg_CPI = mean(CPIAUCSL, na.rm = TRUE))

# Filter S&P 500 data for the years 2000-2023 ensuring one row per year
sp500_filtered <- stocks %>% 
  filter(Year >= 2000 & Year <= 2023) %>% 
  group_by(Year) %>% 
  summarise(Avg_SP500_Close = mean(Close, na.rm = TRUE))

# Ensure only one row per Year and Country in metrics
metrics <- metrics %>% 
  group_by(Year, Country) %>% 
  summarise(across(everything(), mean, na.rm = TRUE), .groups = "drop") %>% 
  mutate(Is_US = ifelse(Country == "United States", "US", "Non-US"))

# Start merged_data with metrics to ensure all countries are included
merged_data <- metrics

# Merge CPI data with all countries
merged_data <- merge(merged_data, cpi_yearly, by = "Year", all.x = TRUE)

# Merge S&P 500 data only with the US rows
merged_data <- merge(merged_data, sp500_filtered, by = "Year", all.x = TRUE)

# Remove S&P 500 values for non-US countries
merged_data <- merged_data %>% 
  mutate(Avg_SP500_Close = ifelse(Is_US == "US", Avg_SP500_Close, NA))

# Correct numerical values using Cost of Living column instead of CPI
merged_data <- merged_data %>%
  mutate(
    Housing_Cost = (Housing_Cost_Percentage / 100) * Cost_of_Living,
    Healthcare_Cost = (Healthcare_Cost_Percentage / 100) * Cost_of_Living,
    Education_Cost = (Education_Cost_Percentage / 100) * Cost_of_Living,
    Transportation_Cost = (Transportation_Cost_Percentage / 100) * Cost_of_Living,
    Tax_Cost = (Tax_Rate / 100) * Cost_of_Living
  )

# Display row counts before and after merging for validation
print(paste("Row count of sp500_filtered:", nrow(sp500_filtered)))
print(paste("Row count of cpi_yearly:", nrow(cpi_yearly)))
print(paste("Row count of metrics:", nrow(metrics)))
print(paste("Row count of merged_data:", nrow(merged_data)))

# Display the merged dataset
print(merged_data)
```
```{r}
head(cpi)
```

##Introduction: 
As retirement

##Setup: Loading Libraries and Data 

##Research Q1 
###Data / Visualization
What is the correlation of S&P 500 data against the average consumer price index average of U.S Cities, from 2010 - 2023? 

The closing price is the cash value of the last transacted price before the market closes. It is used by investors to measure the stock markets performance. The consumer price index contains numbers that measure changes in the prices of goods and services purchased by house holds in US cities every year.These CPI values are calculated as the weighted averages of price changes for all US cities. This data was collected over the duration of 2010 to 2023.
```{r Research Question 1}
merged_data %>%
  ggplot(aes(Avg_SP500_Close, Avg_CPI)) + geom_line() +
   labs(x = "S&P 500 Closing Price",
        y = "Consumer Price Index",
        title = paste("S&P 500 Closing Price vs Consumer Price Index from 2010 to 2023 per year"))
```
After looking at this visualization, one can conclude that as the S&P 500 Closing Price increases, the Consumer Price Index increases as well. If the closing price is increasing, this means the stock market is performing better. Similarly, if the CPI is increasing, this means that the average price of goods purchased in US cities is increasing, meaning that people are spending more and there is economic growth. Ultimately, this trend suggests that strength in stock market performance can lead to economic growth. 
This data was collected from 2010 to 2023 and is monthly. For starters, it is essential to acknowledge that both the trends in the S&P 500 and CPI for US Cities can be affected by the same confounding factors such as inflation, the employment rate, and geopolitical events such as trade restrictions. Furthermore, correlation does not imply causation.
While creating an experiment is not ethically feasible, to improve upon our findings, we could obtain data on some of the confounding variables such as inflation, the employment rate, and other geopolitical and economic factors. We could create visualizations on the trends between these factors and the S&P 500 and CPI. This will give us insight into causal factors. 

##Research Q2
###Data / Visualization
```{r}
ggplot(merged_data, aes(x = Year)) +
  geom_bar(aes(y = Tax_Rate, fill = "Tax Rate"), stat = "identity", position = "dodge", alpha = 0.6) +
  geom_bar(aes(y = Avg_SP500_Close / max(Avg_SP500_Close) * max(Tax_Rate), fill = "S&P 500"), stat = "identity", position = "dodge", alpha = 0.6) +
  geom_smooth(aes(y = Tax_Rate, color = "Tax Rate Trend"), method = "lm", se = FALSE, size = 1.2) +
  geom_smooth(aes(y = Avg_SP500_Close / max(Avg_SP500_Close) * max(Tax_Rate), color = "S&P 500 Trend"), method = "lm", se = FALSE, size = 1.2) +
  scale_fill_manual(values = c("red", "blue")) +
  scale_color_manual(values = c("darkred", "darkblue")) +
  labs(title = "Tax Rate vs S&P 500 Over Time with Trend Lines",
       x = "Year",
       y = "Scaled Values",
       fill = "Legend",
       color = "Trend Lines") +
  theme_minimal()

```


##Research Q3
###Data / Visualization

The correlation between average individual health care cost per year in the United States and the S&P 500 during the years 2000-2023 is a very weak positive relationship.This means there is no strong linear relationship between tehse two metrics. This can be attributed to other sectors like techonological companies or consumer good companies in the S&P have larger impact on the movement of the S&P 500. Even though healthcare costs have been steadily rising in the United States, based on this data, it does not impact the S&P's overall performance.

```{r}
ggplot(merged_data, aes(x = Housing_Cost, y = Avg_SP500_Close)) +
  geom_point(color = "darkred", size = 1.3, alpha = 0.75) +
  labs(title = "Correlation between Healthcare Cost and S&P 500 from 2000 to 2023",
       x = "Average Individual Healthcare Cost (USD)",
       y = "Average S&P 500 Closing Price")
```

##Research Q4
###Data / Visualization
